---
title: "Tutorial for data preparation"
author: Kaixuan Luo
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE)
```

In this tutorial, we prepare input data for the analysis.

Load the package. 
```{r load-package}
library(mapgen)
```

## GWAS summary statistics

We need a data frame of GWAS summary statistics with the following columns:

* chr: chromosome
* pos: position  
* beta: effect size (if you have Odds Ratio, you will need to transform it to log(Odds Ratio))  
* se: standard error (SE)  
* a0: reference allele
* a1: association/effect allele
* snp: SNP ID (e.g. rsID)
* pval: p-value

Let's load a small example GWAS dataset (`test.sumstats.txt.gz`), and
process the data (see below).

We can use the `vroom` package to read the GWAS summary statistics.
```{r load-gwas-sumstats}
library(vroom)
gwas.file <- system.file('extdata', 'test.sumstats.txt.gz', package='mapgen')
sumstats <- vroom(gwas.file, col_names = TRUE, show_col_types = FALSE)
head(as.data.frame(sumstats),3)
```

## Reference data

Fine-mapping requires linkage-disequilibrium (LD) information. 
You could use either reference genotype panel or LD matrices for the LD blocks. 

### LD blocks

We need a data frame defining LD blocks, with four columns: 
chromosome, start, end positions, and the indices of the LD blocks.

We provided the LD blocks from `LDetect` for the 1000 Genomes (1KG) European population (in hg19).
```{r load-LD-blocks}
LD_Blocks <- readRDS(system.file('extdata', 'LD.blocks.EUR.hg19.rds', package='mapgen'))
head(LD_Blocks, 3)
```


### Reference panel

To use the reference genotype panel, we utilize the R package `bigsnpr` to 
read in PLINK files (bed/bim/fam) into R and 
match alleles between GWAS summary statistics and reference genotype panel. 

If you have reference genotypes in PLINK format, you can use [bigsnpr::snp_readBed()](https://privefl.github.io/bigsnpr/reference/snp_readBed.html) 
to read `bed/bim/fam` files into a [bigSNP][bigSNP] object and save as a '.rds' file.
This '.rds' file could be used for downstream analyses when attached with 
\code{bigsnpr::snp_attach()}.

We provided a `bigSNP` object of the reference genotype panel from 
the 1000 Genomes (1KG) European population. If you are in the He lab at UChicago, 
you can load the `bigSNP` object from RCC as below.
```{r load-bigSNP, eval=FALSE, message=FALSE}
library(bigsnpr)
bigSNP <- snp_attach(rdsfile = '/project2/xinhe/1kg/bigsnpr/EUR_variable_1kg.rds')
```

### LD matrices

We also have pre-computed LD matrices of European samples from UK Biobank. 
They can be downloaded [here][UKBB_LD].
If you are at UChicago, 
you can directly access those LD matrices from RCC at
`/project2/mstephens/wcrouse/UKB_LDR_0.1_b37/`.

We create a data frame `region_info`, with filenames of UKBB reference LD matrices and SNP info 
adding to the LD_Blocks.
```{r get-region-info}
region_info <- get_UKBB_region_info(LD_Blocks,
                                    LDREF.dir = "/project2/mstephens/wcrouse/UKB_LDR_0.1_b37", 
                                    prefix = "ukb_b37_0.1")
```

_* You can skip the LD matrices if 
you only need to run enrichment analysis, 
or if you only need to run downstream analysis (e.g. gene mapping) with 
precomputed fine-mapping result._

## Process GWAS summary statistics

We run `process_gwas_sumstats()` to process the summary statistics.
This checks and cleans GWAS summary statistics, 
add locus ID from the `LD_Blocks` if available,
match alleles in GWAS SNPs with the reference panel from
the `bigSNP` object if `bigSNP` is specified, 
or harmonize GWAS SNPs with the reference LD information 
from the precopmuted LD matrices if `region_info` is specified. 

```{r process-gwas-sumstats, eval=FALSE}
gwas.sumstats <- process_gwas_sumstats(sumstats, 
                                       chr='chr', 
                                       pos='position_b37', 
                                       beta='bcac_onco2_beta', 
                                       se='bcac_onco2_se',
                                       a0='a0', 
                                       a1='a1', 
                                       snp='phase3_1kg_id', 
                                       pval='bcac_onco2_P1df_Wald',
                                       LD_Blocks=LD_Blocks,
                                       bigSNP=bigSNP)
```

Check that the summary statistics is processed and has appropriate columns:
```{r include=FALSE}
gwas.sumstats <- readRDS(system.file('extdata', 'test.processed.sumstats.rds', package='mapgen'))
```

```{r print-gwas-sumstats}
head(as.data.frame(gwas.sumstats),3)
```

[enrichment-tutorial]: https://xinhe-lab.github.io/mapgen/articles/enrichment_finemapping_tutorial.html
[finemapping-tutorial]: https://xinhe-lab.github.io/mapgen/articles/enrichment_finemapping_tutorial.html
[gene-mapping-tutorial]: https://xinhe-lab.github.io/mapgen/articles/gene_mapping_tutorial.html
[bigSNP]: https://privefl.github.io/bigsnpr/reference/bigSNP-class.html
