---
title: "Fine-mapping and LD diagnosis with UK Biobank reference LD matrices"
author: Kaixuan Luo
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE)
```

Inconsistencies between GWAS z-scores and the LD reference could 
potentially inflate PIPs from SuSiE finemapping. It would be helpful to 
check potential LD mismatch issue, for example using [DENTIST][DENTIST] or 
the [diagnostic][diagnostic-susie-rss] from `susie_rss`.

In this tutorial, we show an example of performing fine-mapping and
LD mismatch diagnosis (using `susie_rss`) 
with pre-computed LD matrices from UK Biobank reference.

We have pre-computed the LD matrices of European samples from UK Biobank. 
They can be downloaded [here][UKBB_LD].

Load the packages. 
```{r load-packages, eval=TRUE, message=FALSE}
# library(mapgen)
library(tidyverse)
library(ggplot2)
devtools::load_all("/home/kaixuan/projects/mapgen_package/mapgen_LD_test/mapgen/.")
```

Load an example Asthma GWAS summary statistics
```{r load-GWAS}
gwas.file <- '/project2/xinhe/shared_data/mapgen/example_data/GWAS/aoa_v3_gwas_ukbsnps_susie_input.txt.gz'
gwas.sumstats <- vroom::vroom(gwas.file, col_names = TRUE, show_col_types = FALSE)
head(gwas.sumstats)

n = 336210
```

Load LD regions and add filenames of UKBB reference LD matrices and SNP info
```{r}
LD_Blocks <- readRDS(system.file('extdata', 'LD.blocks.EUR.hg19.rds', package='mapgen'))

region_info <- get_UKBB_region_info(LD_Blocks,
                                    LDREF.dir = "/project2/mstephens/wcrouse/UKB_LDR_0.1_b37", 
                                    prefix = "ukb_b37_0.1")
```

Process GWAS summary statistics and add LD block information,
harmonize GWAS summary statistics with LD reference,
flip signs when reverse complement matches and remove strand ambiguous variants
```{r process-GWAS}
gwas.sumstats <- process_gwas_sumstats(gwas.sumstats, 
                                       chr='chr', pos='pos', beta='beta', se='se', 
                                       a0='a0', a1='a1', snp='snp', pval='pval',
                                       LD_Blocks = LD_Blocks,
                                       region_info=region_info, 
                                       strand_flip=TRUE, remove_strand_ambig=TRUE)
```

Select GWAS significant loci (pval < 5e-8)
```{r select-gwas}
if (max(gwas.sumstats$pval) > 1) {
  gwas.sumstats <- gwas.sumstats %>% dplyr::mutate(pval = 10^(-pval))
}

sig.loci <- gwas.sumstats %>% dplyr::filter(pval < 5e-8) %>% dplyr::pull(locus) %>% unique()
cat(length(sig.loci), "significant loci. \n")
```

Fine-mapping an example locus with pre-computed UKBB LD matrices (we use uniform prior in this tutorial),
If we set `save = TRUE`, it saves susie result as well as z-scores and R matrix for each locus, 
which could be used later for diagnosis.

```{r}
locus <- sig.loci[2]
sumstats.locus <- gwas.sumstats[gwas.sumstats$locus == locus, ]

susie.results <- run_finemapping(sumstats.locus, region_info = region_info, priortype = 'uniform', n = n, L = 10,
                                 save = TRUE, outputdir = "./test", outname = "LD_test")
susie.locus.res <- susie.results[[1]]
susieR::susie_plot(susie.locus.res, y='PIP')

susie.locus.sumstats <- merge_susie_sumstats(susie.results, sumstats.locus)

z.locus <- susie.locus.res$z
R.locus <- susie.locus.res$R
```

Inconsistencies between GWAS z-scores and the LD reference could 
potentially inflate PIPs from SuSiE finemapping. It would be helpful to 
check for potential LD mismatch issues, for example using [DENTIST][DENTIST] or 
[susie_rss][diagnostic-susie-rss]. 

Here we perform LD mismatch diagnosis for this example locus using 
[susie_rss][diagnostic-susie-rss]:

```{r, message=FALSE}
condz <- LD_diagnosis_susie_rss(z.locus, R.locus, n)
condz$plot
```

Next, we introduce LD mismatches by flipping alleles for 10 variants (with large effects) 
and see if we can detect those.
```{r}
seed = 1
set.seed(seed)

flip_index <- sample(which(sumstats.locus$zscore > 2), 10)
sumstats.locus.flip <- sumstats.locus
sumstats.locus.flip$zscore[flip_index] <- -sumstats.locus$zscore[flip_index]

cat(length(flip_index), "allele flipped variants:\n", sort(sumstats.locus.flip$snp[flip_index]), "\n")
```

Then, we run fine-mapping including variants with flipped alleles.
```{r}
susie.results <- run_finemapping(sumstats.locus.flip, region_info = region_info, priortype = 'uniform', n = n, L = 10)
susie.locus.res <- susie.results[[1]]
susieR::susie_plot(susie.locus.res, y='PIP')
```

Let's compare observed z-scores vs the expected z-scores.
```{r, message=FALSE}
condz <- LD_diagnosis_susie_rss(sumstats.locus.flip$zscore, R = R.locus, n = n)
# condz$plot
```

Detect possible allele flipped variants (logLR > 2 and abs(z) > 2).
```{r diagnosis}
detected_index <- which(condz$conditional_dist$logLR > 2 & abs(condz$conditional_dist$z) > 2)
cat(sprintf("Detected %d variants with possible allele flipping", length(detected_index)), "\n")
cat("Possible allele switched variants:", sort(sumstats.locus.flip$snp[detected_index]), "\n")

condz$conditional_dist$flipped <- 0
condz$conditional_dist$flipped[flip_index] <- 1

condz$conditional_dist$detected <- 0
condz$conditional_dist$detected[detected_index] <- 1

cat(sprintf("%d out of %d flipped variants detected with logLR > 2 and abs(z) > 2. \n", 
            length(intersect(detected_index, flip_index)), length(flip_index)))

condz$conditional_dist[union(flip_index, detected_index),]
```

```{r diagnosis-plots}
ggplot(condz$conditional_dist, aes(x = condmean, y = z, col = factor(flipped))) +
  geom_point() +
  scale_colour_manual(values = c("0" = "black", "1" = "red")) + 
  labs(x = "Expected value", y = "Observed z scores", color = "Simulated allele flipping") + 
  theme_bw()

ggplot(condz$conditional_dist, aes(x = condmean, y = z, col = factor(detected))) +
  geom_point() +
  scale_colour_manual(values = c("0" = "black", "1" = "red")) + 
  labs(x = "Expected value", y = "Observed z scores", color = "Detected allele flipping") + 
  theme_bw()
```

[DENTIST]: https://doi.org/10.1038/s41467-021-27438-7
[diagnostic-susie-rss]: https://stephenslab.github.io/susieR/articles/susierss_diagnostic.html
[UKBB_LD]: https://uchicago.box.com/s/jqocacd2fulskmhoqnasrknbt59x3xkn
