---
title: "Tutorial for making track plots"
author: Kaixuan Luo
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

## Input data

* Fine-mapping summary statistics.
* Gene annotations.
* Functional annotation data, e.g.: chromatin loops, open chromatin regions,
histone modification data, etc.

Please have the following Bioconductor packages installed:
`GenomicFeatures`, `rtracklayer`, `Gviz`, 
`GenomicInteractions`, `AnnotationDbi`, `org.Hs.eg.db`.

Load packages
```{r load-packages, message=FALSE, warning=FALSE}
library(GenomicFeatures) #  Making and manipulating annotations
library(rtracklayer) # Import annotation data
library(Gviz) # R package used to visualize track plots
library(GenomicInteractions) # visualize HiC loops
library(AnnotationDbi) # match gene ID to gene symbol
library(org.Hs.eg.db) # match gene ID to gene symbol
library(mapgen)
```

If you are in Xin He lab at U Chicago, you can access the example data on RCC:
```{r, eval=TRUE}
trackdata.dir <- "/project2/xinhe/shared_data/mapgen/example_data/trackplot"
```

### Fine-mapping summary statistics 

Here we use the fine-mapping summary statistics from the AFib study.
```{r load-finemapping-res}
finemapstats <- readRDS(system.file("extdata", "AF.finemapping.sumstats.rds", package = "mapgen"))
finemapstats <- process_finemapping_sumstats(finemapstats, 
                                             snp = 'snp', 
                                             chr = 'chr', 
                                             pos = 'pos', 
                                             pip = 'susie_pip', 
                                             pval = 'pval', 
                                             zscore = 'zscore', 
                                             cs = 'cs', 
                                             locus = 'locus',  
                                             pip.thresh = 0)
```

### Gene annotations

You can download gene annotations (GTF file) from [GENCODE][GENCODE], 
and make gene annotations using the `make_genomic_annots()` function with the GTF file. 

```{r, eval=FALSE}
gtf_file <- '/project2/xinhe/shared_data/gencode/gencode.v19.annotation.gtf.gz'
genomic.annots <- make_genomic_annots(gtf_file)
```

We included gene annotations (hg19) in the package, downloaded from 
[GENCODE release 19][GENCODE V19].
```{r load-annotations}
genomic.annots <- readRDS(system.file("extdata", "genomic.annots.hg19.rds", package = "mapgen"))
gene.annots <- genomic.annots$genes
```

### TxDb database

We can build a `TxDb` database (".sqlite") using the GTF file,
and use \code{loadDb()} to load the `TxDb` database.
```{r make-txdb, eval=FALSE}
gtf_file <- '/project2/xinhe/shared_data/gencode/gencode.v19.annotation.gtf.gz'
txdb <- makeTxDbFromGFF(gtf_file, format = "gtf")
saveDb(txdb, "gencode.v19.annotation.gtf.sqlite")
```

If you are in Xin He lab at UChicago, you can access the gene annotations and 
`TxDb` database from RCC. 
```{r, eval=TRUE}
gtf_file <- '/project2/xinhe/shared_data/gencode/gencode.v19.annotation.gtf.gz'
txdb <- loadDb("/project2/xinhe/shared_data/gencode/gencode.v19.annotation.gtf.sqlite")
```

### Genomic annotations

Load promoter-capture HiC (PC-HiC) data from cardiomyocytes (CMs).
```{r load-pcHiC}
pcHiC <- readRDS(system.file("extdata", "pcHiC.CM.gr.rds", package = "mapgen"))
pcHiC <- pcHiC[pcHiC$gene_name %in% gene.annots$gene_name, ] # restrict to protein coding genes
```

Load ABC scores from heart ventricle (from Nasser *et al.* *Nature* 2021).
```{r process_ABC, message=FALSE}
ABC <- data.table::fread(system.file("extdata", "heart_ventricle-ENCODE_ABC.tsv.gz", package = "mapgen"))
ABC <- process_ABC(ABC, full.element = TRUE)
ABC <- ABC[ABC$gene_name %in% gene.annots$gene_name, ] # restrict to protein coding genes
ABC$score <- ABC$score * 100 # scale to visualize the ABC scores
head(ABC, 3)
```

Load H3K27ac and DHS `.bed` files.

```{r load-bed-files}
H3K27ac_peaks <- rtracklayer::import(file.path(trackdata.dir, "H3K27ac.heart.concat.hg19.bed.gz"))
DHS_peaks <- rtracklayer::import(file.path(trackdata.dir, "FetalHeart_E083.DNase.hg19.narrowPeak.bed.gz"))
```

Load ATAC-seq counts data (the data should be in `wig`, `bigWig/bw`, `bedGraph`, or `bam` format)
```{r load-counts-data}
CM_counts <- rtracklayer::import(file.path(trackdata.dir, "Cardiomyocyte.atac.hg19.bedGraph.gz"))
Endo_counts <- rtracklayer::import(file.path(trackdata.dir, "Endothelial.atac.hg19.bedGraph.gz")) 
Fibro_counts <- rtracklayer::import(file.path(trackdata.dir, "Fibroblast.atac.hg19.bedGraph.gz"))
```

### LD reference panel

If you want to visualize r^2 between SNPs, 
we need a reference panel in a `bigSNP` object. 
(we will implement an option that takes LD matrix as input)

If you don't provide the `bigSNP` object, 
the SNPs in the GWAS track will be plotted in the same color. 

```{r, message=FALSE, warning=FALSE}
library(bigsnpr) # loading reference genotype for LD calculation
```

If you are in the He lab at UChicago, you can load the `bigSNP` object 
from the 1000 Genomes (1KG) European population. 
```{r load-bigSNP, eval=TRUE}
bigSNP <- bigsnpr::snp_attach(rdsfile = '/project2/xinhe/1kg/bigsnpr/EUR_variable_1kg.rds')
```

## Make track plots

Plot HCN4 locus in the genomic region "chr15:73610000-73700000"

Highlight SNP "rs7172038"
```{r make-HCN4-trackplot, fig.width=12, fig.height=7}
counts <- list("CM" = CM_counts, "Endo" = Endo_counts, "Fibro" = Fibro_counts)
peaks <- list("H3K27ac" = H3K27ac_peaks, "DHS" = DHS_peaks)
loops <- list("ABC" = ABC)

track_plot(finemapstats,
           region = "chr15:73610000-73700000",
           gene.annots,
           bigSNP = bigSNP,
           txdb = txdb,
           counts = counts,
           peaks = peaks,
           loops = loops,
           genome = "hg19",
           filter_loop_genes = "HCN4",
           highlight_snps = "topSNP", 
           counts.color = c("red", "green", "purple"),
           peaks.color = c("navy", "blue"),
           loops.color = "gray", 
           genelabel.side = "above",
           verbose = TRUE)
```

[GENCODE]: https://www.gencodegenes.org
[GENCODE V19]: https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_19/gencode.v19.annotation.gtf.gz
